### 2016-07-04 case-studies.tex
Final set of major edits.

In this section, we present two case studies that demonstrate the
applicability of our reasoning approach in varied settings, and its
consequent practical utility.

\paragraph{A video view counter.} Our first application is a
Ruby-on-Rails implementation of a counter to count the number of views
a video has garnered on a video-sharing website like
YouTube\footnote{Source code of the applications, experiments, and the
  tools we developed to analyze Ruby can be found at the following
  anonymized url: \url{...}}. The application stores the video count
in a MySQL database, and supports a read operation (a database read
transaction), and an increment operation (a read-increment-write
sequence wrapped in a transaction). The application is expected to
preserve the \emph{monotonicity invariant}, namely that the view count
should never appear to be decreasing. However, the default
\iso{Repeatable Read} isolation level of MySQL leads to anomalous
executions - with 32 concurrent writers performing single increments,
and one reader issuing a constant stream of reads, we observed 12
violations of monotonicity in 10 rounds. We subsequently instantiated
our reasoning framework for an {\sc sc} store with ANSI SQL isolation
levels, and spent 5 man hours determining appropriate isolation levels
needed to enforce the invariant\footnote{The representative \txnimp
  programs we used to formally reason about scenarios described in
  this section can be found in the supplementary.} In particular, we
were able to formally prove that to preserve the invariant, writers
need to execute at \iso{Serializable} isolation, while readers can
execute at \iso{Read Uncommitted}. Repeating the experiments with this
configuration led to no violations, as expected. While executing all
transactions at {\sc ser} also prevents violations, it led to an
additional 113\% increase in the latency of writes on an average (32
concurrent writers; 5 rounds). We repeated the experiments with the
same configuration on Postgres, and found no violations. However,
Postgres's isolation levels~\cite{postgresiso} are stronger than ANSI
SQL's in non-trivial ways. We therefore formalized Postgres'
guarantees and repeated relevant parts of the proof to discover that
\iso{Repeatable Read} is in fact sufficient for writers - an
observation we confirmed through experiments. As usual, using {\sc
  ser} for all transactions also suffices, but on Postgres results in
an additional 47\% increase in write latency.

@@

\paragraph{Microblog.} Our second set of experiments focused on a Twitter-like microblogging
site, also written using Ruby-on-Rails~\cite{railsbook}. One of the
invariants that the application chooses to enforce is referential
integrity between microposts and authors. The application enforces
this invariant entirely in user space by relying on database-backed
transactions (this behavior is typical of Rails

@@

referential integrity. To determine appropriate isolation levels, we
then instantiated our framework for an {\sc sc} store with various
ANSI SQL isolation levels for user deletions and timeline reads, and
obtained a proof that {\sc rc} is sufficient for deletions, whereas
reads require {\sc rr}. The proof is immediately applicable for MySQL,
which implements ANSI SQL isolation levels. Repeating our experiments
for MySQL with this configuration resulted in no violations (as
expectected), while yielding a 38\% reduction in latency than an {\sc
  ser} configuration (96 concurrent clients; 5 rounds).  Since
Postgres's isolation levels are stronger than the corresponding ANSI
levels, \rulelabel{RG-Conseq} rule (Fig.~\ref{fig:rg-rules}) lets us

@@

%  GK: what does the following sentence mean?
%  As a consequence, we avoid the need for semantics re-engineering
%  every time an application is retargeted to a new platform.

>>>>

In this section, we present two case studies that demonstrate the
applicability of our reasoning approach in varied settings, and its
consequent practical utility. Both our case studies target
Ruby-on-Rails applications. To analyze these, and other similar
open-source applications, we developed a symbolic execution engine
called {\sc MagLev} that compiles a Rails program to an abstract
program in an extended version of \txnimp\footnote{The details of {\sc
MagLev} can be found in the supplementary. Its source code, along with
that of our applications and experiments, is available at the
following anonymized url: \url{...}}. The focus of our formal
verification efforts are such \txnimp programs.

\paragraph{A video view counter.} Our first application is a Rails
implementation of a counter to count the number of views a video has
garnered on a video-sharing website like YouTube. The application
stores the video count in a MySQL database, and supports a read
operation (a database read transaction), and an increment operation (a
read-increment-write sequence wrapped in a transaction).  The
application is expected to preserve the \emph{monotonicity invariant},
namely that the view count should never appear to be decreasing.
However, the default \iso{Repeatable Read} isolation level of MySQL
leads to anomalous executions - with 32 concurrent writers performing
single increments, and one reader issuing a constant stream of reads,
we observed 12 violations of monotonicity in 10 rounds. We
subsequently instantiated our reasoning framework for an {\sc sc}
store with ANSI SQL isolation levels, and spent 5 man hours
determining appropriate isolation levels needed to enforce the
invariant. In particular, we were able to formally prove that to
preserve the invariant, writers need to execute at \iso{Serializable}
isolation, while readers can execute at \iso{Read Uncommitted}.
Repeating the experiments with this configuration led to no
violations, as expected. While executing all transactions at {\sc ser}
also prevents violations, it led to an additional 113\% increase in
the latency of writes on an average (32 concurrent writers; 5 rounds).
We repeated the experiments with the same configuration on Postgres,
and found no violations. However, Postgres's isolation
levels~\cite{postgresiso} are stronger than ANSI SQL's in non-trivial
ways. We therefore formalized Postgres' guarantees and repeated
relevant parts of the proof to discover that \iso{Repeatable Read} is
in fact sufficient for writers - an observation we confirmed through
experiments. As usual, using {\sc ser} for all transactions also
suffices, but on Postgres results in an additional 47\% increase in
write latency.

@@

\paragraph{Microblog.} Our second set of experiments focused on a
Twitter-like microblogging application taken from~\cite{railsbook}.
One of the invariants that the application chooses to enforce is
referential integrity between microposts and authors. The application
enforces this invariant entirely in user space by relying on
database-backed transactions (this behavior is typical of Rails

@@

referential integrity (the isolation level was left at the default
{\sc rc} level for experiments). To determine appropriate isolation
levels, we then instantiated our framework for an {\sc sc} store with
various ANSI SQL isolation levels for user deletions and timeline
reads, and obtained a proof that {\sc rc} is sufficient for deletions,
whereas reads require {\sc rr}. The proof is immediately applicable
for MySQL, which implements ANSI SQL isolation levels. Repeating our
experiments for MySQL with this configuration resulted in no
violations (as expectected), while yielding a 38\% reduction in
latency than an {\sc ser} configuration.  Since Postgres's isolation
levels are stronger than the corresponding ANSI levels,
\rulelabel{RG-Conseq} rule (Fig.~\ref{fig:rg-rules}) lets us

@@

### 2016-07-04 motivation.tex
Final set of major edits.

As an alternative, consider the execution of this transaction under
\emph{Read Committed} ({\sc rc}) isolation level, which is weaker than
{\sc ser}\footnote{{\sc rc} is in fact the default isolation level in
Postgres 9.5 and Oracle 11g databases.} An {\sc rc} transaction is
isolated from the writes of uncommitted transactions, and is therefore
free from \emph{dirty reads}~\cite{berenson} that read uncommitted
data. In the current example, {\sc rc} isolation admits the following
two executions (Fig.~\ref{fig:rc-ex}) on a strongly consistent ({\sc
sc}) store, such as a relational database:

\begin{figure}[!h]
\centering
\subcaptionbox {
  {\sc rc} Execution 1
  \label{fig:motiv-eg-1-a}
} [
  0.55\columnwidth
] {
  \includegraphics[scale=0.5]{Figures/motiv-eg-1-a}
}
%\hspace*{0.5in}
\subcaptionbox {
  {\sc rc} Execution 2
  \label{fig:motiv-eg-1-b}
}{
  \includegraphics[scale=0.5]{Figures/motiv-eg-1-b}
}
\caption{\small A possible execution of the program shown in Fig.~\ref{fig:motiv-eg-1} under
  a read committed isolation level.}
\label{fig:rc-ex}
\end{figure}
The figure depicts an execution as a series of read, write, and commit
operations.  The subscript of an operation indicates the transaction
that executes it. In the first execution, transaction \C{Wd1} reads
the current balance (\C{k}) and writes the new balance (\C{k-a1}), but
before it commits transaction \C{Wd2} executes and commits, writing
the new balance (\C{k-a2}). {\sc rc} isolation prevent \C{Wd2} from
witnessing the uncommitted writes of transaction \C{Wd1}.
Subsequently committing \C{Wd1} leads to the loss of \C{Wd2}'s updates
(the so called \emph{lost update anomaly}~\cite{berenson}), resulting
in an incorrect balance of \C{k-a1}. The second execution describes a
similar scenario with \C{Wd1} and \C{Wd2} exchanging their roles.
Clearly, {\sc rc} is an excessively weak isolation level for this
program because it loses the updates of one transaction, resulting in
the violation of the post-condition.  A stronger isolation level that
prevents lost updates is required. \iso{Snapshot Isolation} ({\sc
si})~\cite{berenson} fits this requirement; {\sc si} effectively
serializes transactions that update a shared data object by aborting
and re-executing a transaction if write-write conflicts are detected
during its commit.\footnote{{\sc si} however does not serialize
transactions in the absence of write-write conflicts.} Since {\sc si},
unlike {\sc ser}, does not rely on expensive lock-based concurrency
control, it is also more efficient, making it appropriate for both
\C{Wd} transactions.

@@

\caption{\small $\hbZ$ arrows allowed by {\sc si} (solid) and {\sc rc} (solid
  and dashed) isolation levels for the example in
  Fig.~\ref{fig:motiv-eg-1}. Each arrow represents an $\hbZ$ relationship
  among all operations of `Wd2' and operations in `Wd1' that follow
  the arrow head. Letters R, W and C stand for read, write and commit,
  respectively.}
\label{fig:motiv-eg-1-hb}
\vspace*{-8pt}
\end{figure}

For the rely-guarantee approach to be useful, it has to thus constrain
allowed interference in accordance with the chosen level of isolation
associated with a transaction. The first observation we make is that
we can constrain interference by constraining the definition of the
\emph{happens-before} ($\hbZ$) relation that dictates visibility
actions between transactions, or their respective constituents.  To do
so, we axiomatize the $\hbZ$ relation to capture the interference
characteristics of different isolation levels as their specifications.
Two such specifications are shown\footnote{Note that specifications
presented here are solely for illustration. The actual specifications
described in \S\ref{sec:ansi-isolation} are more nuanced.} in
Fig.~\ref{fig:interference-ex}.  \iso{Read Committed} specification
($\psi_{RC}$) allows an operation $\eta_1$ of $T_1$ to happen before
an operation $\eta_2$ of $T_2$ only if every operation in $T_1$
(including its commit) happens before $\eta_2$ (we let $T_1 \hboar
\eta_2$ denote this). The specification does not require $T_1$ to
execute and commit before all operations of $T_2$, thus allowing it to
interfere in $T_2$. The \iso{Snapshot Isolation} specification
($\psi_{SI}$) requires two transactions that write to the same
variable to be related by happens-before.  This effectively prohibits
interference due to actions in $T_1$ being interleaved in $T_2$ (or,
vise versa).

Fig.~\ref{fig:motiv-eg-1-hb} is a visualization of $\hbZ$ 
from \C{Wd2} to \C{Wd1} allowed by $\psi_{RC}$ and $\psi_{SI}$. All
arrows are legal under $\psi_{RC}$ because in no case does an operation
$\eta_1$ from \C{Wd1} happen-before $\eta_2$ of \C{Wd2} without the commit
of \C{Wd1} also happening before $\eta_2$. $\psi_{SI}$ however disallows
all dashed arrows, because dashed arrows establish $\hbZ$ between actions in \C{Wd2}
and only a subset of the operations in \C{Wd1}. Ligher (blue) arrows with
hollow heads denote $\hbZ$ relationships that do not affect the value
of \C{C} in a way that causes the program to violate its
post-condition.  Darker (black) arrows with solid heads denote $\hbZ$
relationships that lead to the violation of the post-condition. The task of
the reasoning framework is to determine if all $\hbZ$ relationships
allowed by an isolation level lead to the satisfaction of the
post-condition.

@@

\noindent $\phi$ asserts that \C{Wd1} is not yet committed, and that it wrote to
\C{C}, and the value of \C{C} is either \C{k-a1-a2} or \C{k-a1}
depending on whether or not \C{Wd2} is committed. If $\phi$ remains
invariant until \C{Wd1} commits, then the post-condition (\C{C =
k-a1-a2}) can be established easily. However, an interference from
\C{Wd2} at this stage (captured by the last dashed arrow in
Fig.~\ref{fig:motiv-eg-1-hb}) may violate the invariant by writing
$\C{k-a1}$ to \C{C} and committing \C{Wd2}, thus leading to
$\committed(\C{Wd2}) \Rightarrow \C{C=k-a2}$. Fortunately,
\iso{Snapshot Isolation} prevents this interference, and this can be
shown by demonstrating that an interference from \C{Wd2} starting from an
execution state that satisfies $\psi_{SI} \wedge \phi$ leads to an
execution state where neither $\C{Wd2} \hboar \C{Wd1}$ nor $\C{Wd1}
\hboar \C{Wd2}$ holds (the former does not hold because at least one write
from \C{Wd1} has happened before at least one operation of \C{Wd2}, and
the latter does not hold because \C{Wd1} has not yet committed, and \C{Wd2}
has already begun). Since \C{Wd2} also writes to \C{C}, this violates
the $\psi_{SI}$ constraint which we assume to be an invariant. A proof
that the post-condition holds now follows from the contradiction. It is
informative to note that if the invariant is $\psi_{RC}$ instead of
$\psi_{SI}$, we cannot derive a contradiction and we cannot rule out
the interference, which (rightfully) causes  the proof to fail.

>>>>

As an alternative, consider the execution of this transaction under
\emph{Read Committed} ({\sc rc}) isolation level, which is weaker than
{\sc ser}\footnote{{\sc rc} is in fact the default isolation level in
Postgres 9.5 and Oracle 11g databases.} An {\sc rc} transaction is
isolated from the writes of uncommitted transactions, and is therefore
free from \emph{dirty reads}~\cite{berenson} that read uncommitted
data. In the current example, {\sc rc} isolation admits the two
executions shown in Fig.~\ref{fig:rc-ex} on a strongly consistent
({\sc sc}) store, such as a relational database.

\begin{figure}[!h]
\centering
\subcaptionbox {
  {\sc rc} Execution 1
  \label{fig:motiv-eg-1-a}
} [
  0.55\columnwidth
] {
  \includegraphics[scale=0.6]{Figures/motiv-eg-1-a}
}
%\hspace*{0.5in}
\subcaptionbox {
  {\sc rc} Execution 2
  \label{fig:motiv-eg-1-b}
}{
  \includegraphics[scale=0.6]{Figures/motiv-eg-1-b}
}
\caption{\small A possible execution of the program shown in Fig.~\ref{fig:motiv-eg-1} under
  \iso{Read Committed} isolation level.}
\label{fig:rc-ex}
\end{figure}

The figure depicts an execution as a series of read, write, and commit
operations.  The subscript of an operation indicates the transaction
that executes it. In the first execution, transaction \C{Wd1} reads
the current balance (\C{k}) and writes the new balance (\C{k-a1}), but
before it commits transaction \C{Wd2} executes and commits, writing
the new balance (\C{k-a2}). {\sc rc} isolation prevent \C{Wd2} from
witnessing the uncommitted writes of transaction \C{Wd1}.
Subsequently committing \C{Wd1} leads to the loss of \C{Wd2}'s updates
(the so called \emph{lost update anomaly}~\cite{berenson}), resulting
in an incorrect balance of \C{k-a1}. The second execution describes a
similar scenario with \C{Wd1} and \C{Wd2} exchanging their roles.
Clearly, {\sc rc} is an excessively weak isolation level for this
program because it loses the updates of one transaction, resulting in
the violation of the post-condition.  A stronger isolation level that
prevents lost updates is required. \iso{Snapshot Isolation} ({\sc
si})~\cite{berenson} fits this requirement; {\sc si} effectively
serializes transactions that update a shared data object by aborting
and re-executing a transaction if write-write conflicts are detected
during its commit.\footnote{{\sc si} however does not serialize
transactions in the absence of write-write conflicts.} Since {\sc si},
unlike {\sc ser}, does not rely on expensive lock-based concurrency
control, it is also more efficient, making it appropriate for both
\C{Wd} transactions.

@@

\caption{\small $\hbZ$ arrows allowed by {\sc si} (solid) and {\sc rc} (solid
  and dashed) isolation levels for the example in
  Fig.~\ref{fig:motiv-eg-1}. Each arrow represents an $\hbZ$ relationship
  among all operations of \C{Wd2} and operations in \C{Wd1} that follow
  the arrow head. Letters R, W and C stand for read, write and commit,
  respectively.}
\label{fig:motiv-eg-1-hb}
\vspace*{-8pt}
\end{figure}

For the rely-guarantee approach to be useful, it has to thus constrain
allowed interference in accordance with the chosen level of isolation
associated with a transaction. The first observation we make is that
we can constrain interference by constraining the definition of the
\emph{happens-before} ($\hbZ$) relation, which ultimately dictates
what actions become visible to a transaction, or its constituents, and
when. To do so, we axiomatize the $\hbZ$ relation to capture the
interference characteristics of different isolation levels as their
specifications.  Two such specifications are shown\footnote{Note that
specifications presented here are solely for illustration. The actual
specifications described in \S\ref{sec:ansi-isolation} are more
nuanced.} in Fig.~\ref{fig:interference-ex}.  \iso{Read Committed}
specification ($\psi_{RC}$) allows an operation $\eta_1$ of $T_1$ to
happen before an operation $\eta_2$ of $T_2$ only if every operation
in $T_1$ (including its commit) happens before $\eta_2$ (we let $T_1
\hboar \eta_2$ denote this). The specification does not require $T_1$
to execute and commit before all operations of $T_2$, thus allowing it
to interfere in $T_2$. The \iso{Snapshot Isolation} specification
($\psi_{SI}$) requires two transactions that write to the same
variable to be related by happens-before.  This effectively prohibits
interference due to actions in $T_1$ being interleaved in $T_2$ (or,
vise versa).

Fig.~\ref{fig:motiv-eg-1-hb} is a visualization of $\hbZ$ from \C{Wd2}
to \C{Wd1} allowed by $\psi_{RC}$ and $\psi_{SI}$. All arrows are
legal under $\psi_{RC}$ because in no case does an operation $\eta_1$
from \C{Wd1} happen-before $\eta_2$ of \C{Wd2} without the commit of
\C{Wd1} also happening before $\eta_2$. $\psi_{SI}$ however disallows
all dashed arrows, because dashed arrows establish $\hbZ$ between
actions in \C{Wd2} and only a subset of the operations in \C{Wd1}.
Since the result of an operation (\eg a read) depends on what happens
before that operation, $\hbZ$ relation ultimately affects the result
of the program. The solid arrow and the first (from left) dashed arrow
in Fig.~\ref{fig:motiv-eg-1-hb} denote $\hbZ$ relationships that do
not affect the value of \C{C} in a way that causes the program to
violate its post-condition. The last two dashed arrows denote $\hbZ$
relationships that lead to the violation of the post-condition. The
task of the reasoning framework is to determine if all $\hbZ$
relationships allowed by an isolation level lead to the satisfaction
of the post-condition.

@@

\noindent $\phi$ asserts that \C{Wd1} is not yet committed, and that
it wrote to \C{C}, and the value of \C{C} is either \C{k-a1-a2} or
\C{k-a1} depending on whether or not \C{Wd2} is committed. If $\phi$
remains invariant until \C{Wd1} commits, then the post-condition (\C{C
= k-a1-a2}) can be established easily. However, an interference from
\C{Wd2} at this stage (captured by the last dashed arrow in
Fig.~\ref{fig:motiv-eg-1-hb}) may violate the invariant by writing
$\C{k-a1}$ to \C{C} and committing \C{Wd2}, thus leading to
$\committed(\C{Wd2}) \Rightarrow \C{C=k-a2}$. Fortunately,
\iso{Snapshot Isolation} prevents this interference, and this can be
shown by demonstrating that an interference from \C{Wd2} starting from
an execution state that satisfies $\psi_{SI} \wedge \phi$ leads to an
execution state where neither $\C{Wd2} \hboar \C{Wd1}$ nor $\C{Wd1}
\hboar \C{Wd2}$ holds ($\C{Wd2} \hboar \C{Wd1}$ does not hold because
\C{Wd1}'s write to \C{C} clearly happened before \C{Wd2}'s commit, and
$\C{Wd1} \hboar \C{Wd2}$ does not hold because \C{Wd1} has not yet
committed, while \C{Wd2} has already begun). Since \C{Wd2} also writes
to \C{C}, this violates the $\psi_{SI}$ constraint which we assume to
be an invariant. A proof that the post-condition holds now follows
from the contradiction. It is informative to note that if the
invariant is $\psi_{RC}$ instead of $\psi_{SI}$, we cannot derive a
contradiction and we cannot rule out the interference, which
(rightfully) causes  the proof to fail.

### 2016-07-04 opsem.tex
typo fixes mostly
Final set of major edits.

As described in \S\ref{sec:motivation}, the semantics of various
isolation levels can be captured as constraints over the
happens-before ($\hbZ$) relation. $\hbZ$ is however a derived relation
in our model, composed of more fundamental \emph{session order}
($\soZ$) and \emph{visibility} ($\visZ$) relations, which shall
henceforth replace $\hbZ$ in the discussion.  A session order relation
captures the sequential order of operations within a transaction. In
particular, it relates two effects, $\eta_1$ and $\eta_2$, such that
$\txn(\eta_1) = \txn(\eta_2)$ and $\id(\eta_1) < \id(\eta_2)$.  The
semantics assigns monotonically increasing identifiers to effects, as
defined by the $\id(\eta) > \maxId(\A)$ condition of
\rulelabel{E-Aux}.  Evaluation contexts ($\ectx_i$) for
transaction-bound terms are defined so as to enforce a deterministic
sequential order of execution within a transaction, leading to a
deterministic total order among effect ids, which defines the session
order relation. Visibility ($\visZ$) on the other hand relates effects
across concurrent transactions.  Intuitively, $\visZ$ relates $\eta_1$
to $\eta_2$ if and only if $\eta_1$ was \emph{visible} to the
operation that generated $\eta_2$ during its execution, thus effecting
its return value ($\rval(\eta_2)$). For example, a read operation over
\C{X} may pick the value ($\rval$) of the write effect with the highest id
among its visible effects (this is made possible by appropriately
defining $\interp{\cdot}$ in \rulelabel{E-Var}, as we show later).
Thus, the value of a read depends on what write effects it can
witness. An operation can only witness the effects of already
concluded operations, which varies between executions due to the
non-deterministic order of evaluating the parallel composition of
transactions.

>>>>

As described in \S\ref{sec:motivation}, the semantics of various
isolation levels can be captured as constraints over the
happens-before ($\hbZ$) relation. $\hbZ$ is however a derived relation
in our model, composed of more fundamental \emph{session order}
($\soZ$) and \emph{visibility} ($\visZ$) relations. In particular,
$\hbZ \,=\, (\soZ \cup \visZ)^{+}$. Unlike $\hbZ$, $\visZ$ (defined
below) is not a transitive relation, hence lets us capture finer
properties than $\hbZ$. We therefore let $\visZ$ replace $\hbZ$ in our
discussion henceforth. The session order relation captures the
sequential order of operations within a transaction. In particular, it
relates two effects, $\eta_1$ and $\eta_2$, such that $\txn(\eta_1) =
\txn(\eta_2)$ and $\id(\eta_1) < \id(\eta_2)$.  The semantics assigns
monotonically increasing identifiers to effects, as defined by the
$\id(\eta) > \maxId(\A)$ condition of \rulelabel{E-Aux}. Evaluation
contexts ($\ectx_i$) for transaction-bound terms are defined so as to
enforce a deterministic sequential order of execution within a
transaction, leading to a deterministic total order among effect ids,
which defines the session order relation. Visibility ($\visZ$) on the
other hand relates effects across concurrent transactions.
Intuitively, $\visZ$ relates $\eta_1$ to $\eta_2$ if and only if
$\eta_1$ was \emph{visible} to the operation that generated $\eta_2$
during its execution, thus effecting its return value
($\rval(\eta_2)$). For example, a read operation over \C{X} may pick
the value ($\rval$) of the write effect with the highest id among its
visible effects (this is made possible by appropriately defining
$\interp{\cdot}$ in \rulelabel{E-Var}, as we show later). Thus, the
value of a read depends on what write effects it can witness. An
operation can only witness the effects of already concluded
operations, which varies between executions due to the
non-deterministic order of evaluating the parallel composition of
transactions.
